---
title: "RepData"
author: "Alex MacCalman"
date: "4/23/2020"
output: md_document
---

```{r setup, include=FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(ggplot2)
```
# Activity Monitoring Analysis  
This report analyzes the activity data downloaded from [this link.](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip)

## Loading and preprocessing the data  
```{r}
x <- read.csv("./activity.csv", header = TRUE)
x$date <- as.Date.factor(x$date)
```
## What is the mean total number of steps taken per day?  
Calculate the total number of steps taken per day.
```{r}
groupx <- group_by(x, date)
total <- summarise(groupx, total = sum(steps)) %>% na.omit()
```
Make a histogram of the toal number of steps taken each day.  
```{r totalhist}
ggplot(total, aes(x=total)) + geom_histogram(bins = 25) + labs(x = "Total Steps per Day", y = "Count")
```

Calculate the mean of the total number of steps taken per day.  
```{r}
mean(total$total, na.rm = TRUE)
```
Calculate the median of the total number of steps taken per day.  
```{r}
median(total$total)
```
## What is the average daily activity pattern?  
Group data by interval and plot times series  
```{r avgtime}
groupI <- group_by(x, interval)
interval <- summarise(groupI, Mean = mean(steps, na.rm = TRUE))
ggplot(interval, aes(x = interval, y = Mean)) + geom_line() + labs(x = "5-minute interval", y = "Average Number of Steps Taken")
```

Find the 5-minute interval, on average across all the days in the dataset that contains the maximum number of steps.  
```{r}
 interval[which.max(interval$Mean),]
```
## Imputing missing values  
Calculate the total number of missing values in the dataset. 
```{r}
missingX <- is.na(x$steps)
sum(missingX)
```
Create a lookup table with a key as the date and value as the mean for each date.  
```{r}
meanStep <- summarise(groupx, mean = mean(steps, na.rm = TRUE))
get <- meanStep$mean
names(get) <- as.character(meanStep$date)
```
Create a new data set by replacing the missing values with the mean of each date. Use a for loop to replace the missing values using the look up table.  
```{r}
xnew <- x
for (i in 1:nrow(xnew)) {
  if (is.na(xnew[i,1])) {
    xnew[i,1] <- get[as.character(xnew[i,2])]
  }
}
```
Calculate the total number of steps taken per day with the new data set.  
```{r}
groupxnew <- group_by(xnew, date)
totalnew <- summarise(groupxnew, totalnew = sum(steps)) %>% na.omit()
```
 Make a histogram of the toal number of steps taken each day with the new data set.  
```{r imputedtotals}
ggplot(totalnew, aes(x=totalnew)) + geom_histogram(bins = 25) + labs(x = "Total Steps per Day", y = "Count",title = "Totals with missing values imputed as mean of each day")
```

Calculate the mean of the total number of steps taken per day with the new data set.  
```{r}
mean(totalnew$totalnew)
```

Calculate the median of the total number of steps taken per day with the new data set.  
```{r}
median(totalnew$totalnew)
```

## Are there differences in activity patterns between weekdays and weekends?  

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day.
```{r}
xnew$weekdays <- weekdays(xnew$date)
xnew$day <- ifelse(xnew$weekdays == "Saturday" | xnew$weekdays == "Sunday", "Weekend", "Weekday")
```

```{r compare}
groupInew <- group_by(xnew, interval, day)
intervalnew <- summarise(groupInew, Mean = mean(steps, na.rm = TRUE))
ggplot(intervalnew, aes(x = interval, y = Mean)) + geom_line() + facet_grid(day~.) +labs(x = "5-minute interval", y = "Average Number of Steps Taken", title = "Comparison of intervals betwwen weekdays and weekedns")
```
